\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, shapes.geometric}

\title{Matrix State Resolution v2.1.1: \\ A Surgical Fix for Invite Locks and Ban Evasion}
\author{Shane Jaroch \\ \texttt{chown\_tee@proton.me}}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Matrix State Resolution is the core algorithm for achieving decentralized consensus across federated homeservers. Over its evolution, architectural decisions in the resolution algorithm have inadvertently exposed edge-case vulnerabilities. State Resolution v2 (V2.0) suffered from the "Invite Lock", where aggressive supplemental merges overrode administrative locks. State Resolution v2.1 (MSC4297) solved this by isolating the merge to Power Levels, but accidentally enabled Concurrent Ban Evasion. We present State Resolution v2.1.1, which surgically expands the supplemental merge to include Authoritative Memberships (Bans), achieving the mathematical sweet spot that protects against all known state anomalies.
\end{abstract}

\section{Introduction}
State Resolution is the process by which multiple homeservers in a Matrix room reach a common view of the room state despite participating in a distributed, asynchronous Directed Acyclic Graph (DAG).

\section{The Invite Lock in V2.0}
In V2.0, the "Supplemental Merge" aggressively overlaid all state events, including \texttt{m.room.join\_rules}, into the auth context during resolution.
If an Administrator locked a room to "invite-only", historical \texttt{join} events on slower forks would be erroneously evaluated against the newly merged "invite" rules rather than their local "public" rules. This caused legitimate, historical joins to be rejected during resolution, permanently locking out users.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    node distance=1.5cm and 2cm,
    event/.style={rectangle, draw, rounded corners, minimum width=2cm, minimum height=1cm, align=center},
    auth/.style={->, dashed, thick},
    prev/.style={->, thick}
]

% Nodes
\node[event] (create) {\texttt{\$create}};
\node[event, right=of create] (pl) {\texttt{\$pl}};
\node[event, right=of pl] (public) {\texttt{\$public}};

\node[event, above right=of public, fill=red!10] (invite) {\texttt{\$admin\_lock} \\ (Invite)};
\node[event, below right=of public, fill=green!10] (join) {\texttt{\$hist\_join} \\ (Join)};

% Edges
\draw[auth] (pl) -- (create);
\draw[auth] (public) -- (pl);
\draw[auth] (invite) -- (public);
\draw[auth] (join) -- (public);

\end{tikzpicture}
\caption{The Invite Lock: V2.0 supplements the \texttt{\$admin\_lock} event into the auth overlay, causing the independent \texttt{\$hist\_join} on the slower fork to be evaluated against the "invite" rules instead of its local "public" rules, locking the user out.}
\label{fig:invite_lock}
\end{figure}

\section{Concurrent Ban Evasion in V2.1}
To fix the Invite Lock, Matrix State Resolution v2.1 (MSC4297) strictly isolated the supplemental merge to only include \texttt{m.room.power\_levels}. While this successfully protected \texttt{join\_rules}, it excluded Membership events (Bans).
If an Administrator bans a malicious user, but the malicious user concurrently submits a state event (e.g., a room name change) on a separate fork, V2.1 evaluates the malicious event against the user's local auth chain (which lacks the ban). Because the ban is not supplemented, the malicious event is accepted into the final resolved state.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    node distance=1.5cm and 2cm,
    event/.style={rectangle, draw, rounded corners, minimum width=2.5cm, minimum height=1cm, align=center},
    auth/.style={->, dashed, thick}
]

\node[event] (pl) {\texttt{\$pl} \\ Bob: 50};
\node[event, right=of pl] (join) {\texttt{\$bob\_join}};

\node[event, above right=of join, fill=red!10] (ban) {\texttt{\$alice\_bans} \\ \texttt{\_bob}};
\node[event, below right=of join, fill=orange!10] (name) {\texttt{\$bob\_name} \\ \texttt{\_change}};

\draw[auth] (join) -- (pl);
\draw[auth] (ban) -- (join);
\draw[auth] (name) -- (join);
\draw[auth] (ban) to[bend right=20] (pl);
\draw[auth] (name) to[bend left=20] (pl);

\end{tikzpicture}
\caption{Concurrent Ban Evasion: Under V2.1, the supplemental merge only includes Power Levels (\texttt{\$pl}). Alice's ban is ignored, so Bob's name change is validated against his local auth chain (which considers him joined with PL 50), allowing the evasion.}
\label{fig:ban_evasion}
\end{figure}

\section{State Resolution v2.1.1: The Surgical Merge}
State Resolution v2.1.1 addresses the vulnerabilities of both previous iterations by expanding the v2.1 supplemental merge to include \textbf{Authoritative Memberships}.
By supplementing \texttt{m.room.power\_levels} and \texttt{m.room.member} (specifically when the membership is a "ban" or "leave" and the sender differs from the state key), the consensus Ban successfully overlays onto concurrent malicious events, forcing the iterative auth check to rightfully reject them. Meanwhile, \texttt{m.room.join\_rules} remains cleanly isolated, ensuring the DAG can heal from splits without suffering from Invite Locks.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    node distance=1.5cm and 2cm,
    event/.style={rectangle, draw, rounded corners, minimum width=2.5cm, minimum height=1cm, align=center},
    auth/.style={->, dashed, thick}
]

\node[event, fill=blue!10] (v211) {\textbf{v2.1.1 Auth Overlay}};
\node[event, right=of v211, fill=green!10] (pl) {\texttt{\$pl} (Included)};
\node[event, below=of pl, fill=green!10] (ban) {\texttt{\$ban} (Included)};
\node[event, below=of v211, fill=red!10] (invite) {\texttt{\$admin\_lock} (Excluded)};

\draw[->, thick] (v211) -- (pl);
\draw[->, thick] (v211) |- (ban);
\draw[->, thick, dotted] (v211) -- (invite) node[midway, sloped, above] {Isolated};

\end{tikzpicture}
\caption{State Resolution v2.1.1: The mathematical sweet spot. Power levels and Authoritative Memberships (Bans) are merged into the overlay to defeat evasion attacks, while \texttt{join\_rules} are cleanly isolated to preserve historical DAG integrity.}
\label{fig:v211_merge}
\end{figure}

\section{Defeating Demotion Evasion}
A separate vector, Demotion Evasion, occurs when a demoted administrator attempts to bypass their demotion by referencing an ancient Power Level event in their 1-hop \texttt{auth\_events} chain. Both v2.1 and v2.1.1 inherently defeat this attack because the authoritative demotion is a Power Level event, which is supplemented into the auth overlay and immediately invalidates the malicious event's credentials.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    node distance=1.5cm and 2cm,
    event/.style={rectangle, draw, rounded corners, minimum width=2.5cm, minimum height=1cm, align=center},
    auth/.style={->, dashed, thick}
]

\node[event] (ancient) {\texttt{\$ancient\_pl} \\ Eve: 100};
\node[event, right=of ancient, fill=red!10] (demote) {\texttt{\$pl\_demote} \\ Eve: 0};
\node[event, below right=of ancient, fill=orange!10] (attack) {\texttt{\$eve\_attack} \\ (Requires 50)};

\draw[auth] (demote) -- (ancient);
\draw[auth] (attack) -- (ancient) node[midway, below, sloped] {Bypasses demotion};

\end{tikzpicture}
\caption{Demotion Evasion: Eve maliciously points her event's \texttt{auth\_events} directly to her ancient PL 100 event. Both v2.1 and v2.1.1 defeat this because \texttt{\$pl\_demote} is supplemented into the auth overlay, superseding the ancient PL.}
\label{fig:demotion_evasion}
\end{figure}

\section{Conclusion}
By carefully balancing the supplemental merge, State Resolution v2.1.1 provides a mathematically secure consensus algorithm that protects against both historical lockouts and concurrent evasion attacks.

\end{document}
